Variance reduction for Markov chains via martingale representations

### 2020-04-25 aap/Revision/vr-mcmc-revision-new-new.tex
update

Now, from the H\"older inequality, we obtain (with $\|X\|_p = (\mathsf{E}X^p)^{\frac{1}{p}}$)
\begin{align*}
\mathsf{E}\left[\sum_{j=1}^{q}\prod_{k=l}^{j}\alpha_{k}\left( \sum_{p=j+1}^{q+1}\prod_{k=j+1}^{p}\alpha_{k}\right)^{2}\right] & \leq\sum_{j=1}^{q}\left\Vert \prod_{k=1}^{j}\alpha_{k}\right\Vert _{2}\left\Vert \sum_{p=j+1}^{q+1} \prod_{k=j+1}^{p}\alpha_{k}\right\Vert _{4}^{2}\\
 & \leq\sum_{j=1}^{q} \left\Vert \prod_{k=1}^{j}\alpha_{k}\right\Vert _{2}\left(\sum_{p=j+1}^{q+1}\left\Vert \prod_{k=j+1}^{p}\alpha_{k}\right\Vert _{4}\right)^{2}.
\end{align*}
Now using the fact that $\prod\limits_{k=j+1}^{p}\alpha_{k} \leq \exp\left(-\sum\limits_{k=j+1}^{p}\gamma\mu'(X_{k-1}^x)\right),$
we get
\[
\mathsf{E}\left[\sum_{j=1}^{q}\prod_{k=1}^{j}\alpha_{k}\left(\sum_{p=j+1}^{q+1}\prod_{k=j+1}^{p}\alpha_{k}\right)^{2}\right]\leq\sum_{j=1}^{q} \zeta^{1/2}_{1,j}(2) \left(\sum_{p=j+1}^{q+1} \zeta^{1/4}_{j+1,p}(4)\right)^{2},
\]
where we denote
\[
\zeta_{l,j}(u)=\mathsf{E}\left[\rme^{-u\gamma\sum_{k=l}^{j}\mu'(X^x_{k-1})}\right],\quad u>0.
\]
Note that $\mu^{\prime}(x)$ is bounded due  our assumptions and from \cite[Theorem~1]{delyon1999small} and Lemma~\ref{lem:v_ergodicity} in Appendix~\ref{sec:appendix} it follows that
\[
\mathsf{E}\left[\rme^{-u\gamma\left(\sum_{k=l}^{j}[\mu'(X^x_{k-1})-\pi_\gamma(\mu')]\right)}\right]\leq C_1 \rme^{\varkappa n\gamma^2}(1+V(x)),\quad u\gamma\leq \gamma_0,
\]
for some constants $\gamma_0$ and $C_1,\varkappa>0$. Then we have
\[
\zeta_{l,j}(u)=\mathsf{E}\left[\rme^{-s\gamma\sum_{k=l}^{j}\mu'(X^x_{k-1})}\right]\leq  C_1 \rme^{\varkappa n\gamma^2}(1+V(x)) \rme^{-s\gamma (j-l+1)\pi_\gamma(\mu')}.
\]
Furthermore, since $\mu\pi',\mu'\pi\in L^{1}(\mathbb{R})$ and $\pi^{\prime}(x) = -\frac{1}{2}\pi(x)\mu(x)$, we have
\[
\pi(\mu')=\int\mu'(x)\pi(x)\,dx=-\int\mu(x)\pi'(x)\,dx=\frac{1}{2}\int\mu^{2}(x)\pi(x)\,dx>0
\]
Note also that \(\pi_\gamma(\mu^{\prime}) \geq \pi_\gamma(\mu^{\prime}) - C_2\sqrt{\gamma}\)
yielding the bound
\[
\zeta_{l,j}(s) \leq  C_1 \rme^{\varkappa n\gamma^2}(1+V(x)) \rme^{-s\gamma(j-l+1)\alpha}
\]
where $ \alpha = \frac{1}{2}\int \mu^2(x)\pi(x)\,dx - C_1\sqrt{\gamma} > 0$ for sufficiently small $\gamma$. Hence we obtain
\begin{align*}
\sum_{p=j+1}^{q+1} \zeta^{1/4}_{j+1,p}(4) &\leq (C_1 \rme^{\varkappa n\gamma^2}(1+V(x)))^{1/4}\sum_{p=j+1}^{q+1} \rme^{-\gamma(p-j)\alpha}
\\
&\leq  (C_1 \rme^{\varkappa n\gamma^2}(1+V(x)))^{1/4}\frac{1}{1- \rme^{- \gamma\alpha}}
\end{align*}
and
\begin{align*}
\sum_{j=1}^{q} \zeta^{1/2}_{1,j}(2) \leq (C_1 \rme^{\varkappa n\gamma^2}(1+V(x)))^{1/2}\sum_{j=0}^{q} \rme^{-\gamma j\alpha}
\\
\leq  (C_1 \rme^{\varkappa n\gamma^2}(1+V(x)))^{1/2}\frac{ 1}{1- \rme^{- \gamma\alpha}}.
\end{align*}
Thus the final bound follows:
\begin{align*}
\mathsf{Var}\left[\sum_{p=1}^{q}\partial_{y_{1}}^{K}f\left(X^x_{p}\right)\right] \leq \gamma^{-1} C_K \rme^{\varkappa n\gamma^2} (1+V(x))
\end{align*}
 with $C_K$  depending neither on $q$ nor on $\gamma$. The proof is completed.
\end{proof}

@@

\begin{lem} Let $\mu(x) = \mu_{0}(x) + \Delta(x)$, where $\mu_0^{\prime}(x) <-m$ for some $m > 0$, $\mu_0(x) \in Lip(L)$ and suppose that $\Delta(x)$ is bounded. Then

@@

\PE_x \rme^{u\gamma\sum\limits_{k=N+1}^{N+n}f(X_k)} \leq C\exp\left\{n\gamma\left(\frac{8u^2\|f\|^2_{\text{Lip}}\left(1+\frac{N}{n}\right)(m+L)^2}{m^2L^2} + M^2 + C_1u\sqrt{\gamma} +  u\pi(f)\right)\right\}

@@

with constants $C$ and $C_1$ depending only on $m,L$.

@@

& = & \PE_x \exp\left\{u\gamma\sum_{k=N+1}^{N+n}f(Y_k) + \sum_{k=1}^{n}\left[(Y_k-Y_{k-1})\Delta(Y_{k-1}) - \gamma\mu_0(Y_{k-1})\Delta(Y_{k-1}) - \frac{\gamma}{2}\Delta^2(Y_{k-1})\right]\right\} = \\
 &=& \PE_x \exp\left\{u\gamma\sum_{k=N+1}^{N+n}f(Y_k) + \sum_{k=1}^{n}\left[\sqrt{\gamma}\Delta(Y_{k-1})\xi_{k} - \frac{\gamma}{2}\Delta^2(Y_{k-1})\right]\right\} \leq \\
 & \leq & \left(\PE_x \exp\left\{2\sqrt{\gamma}\sum_{k=N+1}^{N+n}\Delta(Y_{k-1})\xi_{k}\right\}\right)^{1/2} \left(\PE \exp\left\{2u\gamma\sum_{k=1}^{n}f(Y_k)\right\}\right)^{1/2}

@@

\PE \exp\left\{2u\gamma\sum_{k=N+1}^{N+n}\left(f(Y_k) - \PE_xf(X_k)\right)\right\} \exp\left\{2u\gamma\sum_{k=N+1}^{N+n}\PE_x f(Y_k)\right\} 

@@

Now we can apply Lemma~\ref{lem:exp_strongly_convex} to the first expectation with $\lambda = 2nu\gamma$, yielding the bound

@@

\PE \exp\left\{2u\gamma\sum_{k=N+1}^{N+n}\left(f(Y_k) - \PE_xf(X_k)\right)\right\} \leq \exp\left\{\frac{16u^2 n\gamma \|f\|^2_{\text{Lip}}\left(1+\frac{N}{n}\right)(m+L)^2}{m^2L^2}\right\}

@@

\PE \exp\left\{2u\gamma\sum_{k=N+1}^{N+n}f(Y_k)\right\} \leq \exp\left\{\frac{16u^2 n\gamma \|f\|^2_{\text{Lip}}\left(1+\frac{N}{n}\right)(m+L)^2}{m^2L^2} + 2u\gamma \frac{m+L}{mL}\|f\|_{\text{Lip}}S(x) + 2n\gamma u(\pi(f) + C_1\sqrt{\gamma})\right\}

>>>>

Now, from the H\"older inequality, we obtain (with $\|X\|_p = (\mathsf{E}X^p)^{\frac{1}{p}}$)
\begin{align*}
\mathsf{E}\left[\sum_{j=1}^{q}\prod_{k=l}^{j}\alpha_{k}\left( \sum_{p=j+1}^{q+1}\prod_{k=j+1}^{p}\alpha_{k}\right)^{2}\right] & \leq\sum_{j=1}^{q}\left\Vert \prod_{k=1}^{j}\alpha_{k}\right\Vert _{2}\left\Vert \sum_{p=j+1}^{q+1} \prod_{k=j+1}^{p}\alpha_{k}\right\Vert _{4}^{2}\\
 & \leq\sum_{j=1}^{q} \left\Vert \prod_{k=1}^{j}\alpha_{k}\right\Vert _{2}\left(\sum_{p=j+1}^{q+1}\left\Vert \prod_{k=j+1}^{p}\alpha_{k}\right\Vert _{4}\right)^{2}.
\end{align*}
Now using the fact that $\prod\limits_{k=j+1}^{p}\alpha_{k} \leq \exp\left(-\sum\limits_{k=j+1}^{p}\gamma\mu'(X_{k-1}^x)\right),$
we get
\[
\mathsf{E}\left[\sum_{j=1}^{q}\prod_{k=1}^{j}\alpha_{k}\left(\sum_{p=j+1}^{q+1}\prod_{k=j+1}^{p}\alpha_{k}\right)^{2}\right]\leq\sum_{j=1}^{q} \zeta^{1/2}_{1,j}(2) \left(\sum_{p=j+1}^{q+1} \zeta^{1/4}_{j+1,p}(4)\right)^{2},
\]
where we denote
\[
\zeta_{l,j}(u)=\mathsf{E}\left[\rme^{-u\gamma\sum_{k=l}^{j}\mu'(X^x_{k-1})}\right],\quad u>0.
\]
Note that $\mu^{\prime}(x)$ is bounded and Lipschitz due  our assumptions and from Lemma~\ref{lem:exp_bound} in Appendix~\ref{sec:appendix} it follows that
\begin{align*}
\zeta_{l,j}(u) \leq \begin{cases} C(x)\exp\left\{(j-l)\gamma\left(4u^2\kappa^{-2} \|f\|^2_{\text{Lip}} + M^2 + C_1u\sqrt{\gamma} -  u\pi(\mu')\right)\right\}, \quad \gamma(j-l) \geq C_0 := \left(\frac{4}{m+L} + \kappa^{-1}\right) \\
\exp\left\{uC_0\max\limits_{y}|\mu'(y)|\right\}, \quad 0 \leq \gamma(j-l) \leq C_0	
\end{cases}
\end{align*}
for constants $C(x)$ and $C_1$ which could be made precise. Furthermore, since $\mu\pi',\mu'\pi\in L^{1}(\mathbb{R})$ and $\pi^{\prime}(x) = -\frac{1}{2}\pi(x)\mu(x)$, we have
\[
\pi(\mu')=\int\mu'(x)\pi(x)\,dx=-\int\mu(x)\pi'(x)\,dx=\frac{1}{2}\int\mu^{2}(x)\pi(x)\,dx>0
\]
Hence the following bound follows
\begin{align*}
\zeta_{l,j}(u) \leq \begin{cases} C(x)\exp\left\{-\gamma(j-l)\alpha\right\}, \quad \gamma(j-l) \geq C_0 \\
\exp\left\{uC_0\max\limits_{y}|\mu'(y)|\right\}, \quad 0 \leq \gamma(j-l) \leq C_0	
\end{cases}
\end{align*}
with the quantity $ \alpha = \min(...,...) > 0$. Hence we obtain
\begin{align*}
\sum_{p=j+1}^{q+1} \zeta^{1/4}_{j+1,p}(4) &\leq \frac{C_0}{\gamma}\rme^{uC_0\max\limits_{y}|\mu'(y)|} + C(x)\sum_{p=j+1}^{\infty} \rme^{-\gamma(p-j)\alpha}
\leq \frac{C_0}{\gamma}\rme^{uC_0\max\limits_{y}|\mu'(y)|} + \frac{C(x)}{1-\rme^{- \gamma\alpha}}
\end{align*}
and
\begin{align*}
\sum_{j=1}^{q} \zeta^{1/2}_{1,j}(2) \leq \frac{C_0}{\gamma}\rme^{uC_0\max\limits_{y}|\mu'(y)|} + \frac{C(x)}{1-\rme^{- \gamma\alpha}}
\end{align*}
Thus the final bound follows:
\begin{align*}
\mathsf{Var}\left[\sum_{p=1}^{q}\partial_{y_{1}}^{K}f\left(X^x_{p}\right)\right] \leq \gamma^{-1} C_K C(x)
\end{align*}
 with $C_K$  depending neither on $q$ nor on $\gamma$. The proof is completed.
\end{proof}

@@

\label{lem:exp_bound}
Let $(X_p)_{p \geq 0}$ be the ULA-generated Markov chain of the form
X_{p+1} = X_p + \gamma \mu(X_p) + \sqrt{\gamma}\xi_{p+1}, \xi_{p+1} \sim \mathcal{N}(0,1)
 with $0 < \gamma < \frac{4}{m+L}, \quad \mu(x) = \mu_{0}(x) + \Delta(x)$, where $\mu_0^{\prime}(x) <-m$ for some $m > 0$, $\mu_0(x) \in Lip(L)$. Suppose that there exists $M > 0,$ such that $\left|\Delta(x)\right| \leq M$.
 Then for any Lipschitz function $f$, $\forall N $

@@

\PE_x \rme^{u\gamma\sum\limits_{k=N+1}^{N+n}f(X_k)} \leq C\exp\left\{n\gamma\left(2u^2\kappa^{-2} \|f\|^2_{\text{Lip}}\left(1+\frac{1}{n\gamma}\left(\frac{4}{m+L} + \kappa^{-1}\right)\right) + M^2 + C_1u\sqrt{\gamma} +  u\pi(f)\right)\right\}

@@

with positive constants $C < \exp\left\{\frac{4u\|f\|_{\text{Lip}}}{mL}\int_{\rset}|x-y|\,\pi_{\gamma}(dy)\right\}$ and $C_1$ depending only on $m,L$.

@@

& = & \PE_x \exp\left\{u\gamma\sum_{k=N+1}^{N+n}f(Y_k) + \sum_{k=N+1}^{N+n}\left[(Y_k-Y_{k-1})\Delta(Y_{k-1}) - \gamma\mu_0(Y_{k-1})\Delta(Y_{k-1}) - \frac{\gamma}{2}\Delta^2(Y_{k-1})\right]\right\} = \\
 &=& \PE_x \exp\left\{u\gamma\sum_{k=N+1}^{N+n}f(Y_k) + \sum_{k=N+1}^{N+n}\left[\sqrt{\gamma}\Delta(Y_{k-1})\xi_{k} - \frac{\gamma}{2}\Delta^2(Y_{k-1})\right]\right\} \leq \\
 & \leq & \left(\PE_x \exp\left\{2\sqrt{\gamma}\sum_{k=N+1}^{N+n}\Delta(Y_{k-1})\xi_{k}\right\}\right)^{1/2} \left(\PE \exp\left\{2u\gamma\sum_{k=N+1}^{N+n}f(Y_k)\right\}\right)^{1/2}

@@

\PE \exp\left\{2u\gamma\sum_{k=N+1}^{N+n}\left(f(Y_k) - \PE_xf(Y_k)\right)\right\} \exp\left\{2u\gamma\sum_{k=N+1}^{N+n}\PE_x f(Y_k)\right\} 

@@

Now we can apply \cite[Proposition 32]{dm2016HighDim} to the first expectation with $\lambda = 2nu\gamma$, yielding the bound

@@

\PE \exp\left\{2u\gamma\sum_{k=N+1}^{N+n}\left(f(Y_k) - \PE_xf(Y_k)\right)\right\} \leq \exp\left\{4u^2\kappa^{-2} n\gamma \|f\|^2_{\text{Lip}}\left(1+\frac{1}{n\gamma}\left(\frac{4}{m+L} + \kappa^{-1}\right)\right)\right\}

@@

\PE \exp\left\{2u\gamma\sum_{k=N+1}^{N+n}f(Y_k)\right\} \leq \exp\left\{4u^2\kappa^{-2} n\gamma \|f\|^2_{\text{Lip}}\left(1+\frac{1}{n\gamma}\left(\frac{4}{m+L} + \kappa^{-1}\right)\right) + 2u\gamma \frac{m+L}{mL}\|f\|_{\text{Lip}}S(x) + 2n\gamma u(\pi(f) + C_1\sqrt{\gamma})\right\}
